---
title: "Benchmarking Differential Abundance Methods"
author: "Adrien Taudière"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Benchmarking Differential Abundance Methods}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
 collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 7
)
```

## Introduction

This vignette provides a comprehensive benchmark of differential abundance (DA)
methods for microbiome data, inspired by the
[benchdamic](https://bioconductor.org/packages/benchdamic/) package approach.
We compare seven methods:

| Method | Package | Approach |
|--------|---------|----------|
| **ALDEx2** | ALDEx2 | CLR transformation + Monte Carlo sampling from Dirichlet |
| **ANCOM-BC2** | ANCOMBC | Linear regression with bias correction |
| **MaAsLin3** | maaslin3 | Generalized linear models (abundance + prevalence) |
| **DESeq2** | DESeq2 | Negative binomial GLM with size factor normalization |
| **edgeR** | edgeR | Negative binomial GLM with TMM normalization |
| **limma-voom** | limma | Linear models on voom-transformed counts |
| **radEmu** | radEmu | Robust estimation for relative abundance data |

The benchmark evaluates:

1. **Method concordance**: Agreement between methods on significant taxa
2. **Effect size correlation**: Consistency of effect size estimates
3. **Sensitivity/Specificity trade-off**: Using simulated spike-in data
4. **Computational considerations**: Runtime and ease of use

## Setup

```{r load_packages}
library(comparpq)
library(phyloseq)
library(ggplot2)
library(dplyr)
library(tidyr)
library(tibble)
library(patchwork)

# For DA methods
library(DESeq2)
library(edgeR)
library(limma)
library(radEmu)
library(ALDEx2)
library(BiocParallel)

```

## Helper Functions

We define wrapper functions to standardize output across methods.

```{r helper_functions}
#' Run DESeq2 on phyloseq object
#' @param physeq A phyloseq object
#' @param formula Formula for the design (e.g., "~ condition")
#' @param contrast Character vector for contrast (e.g., c("condition", "B", "A"))
#' @param nclusters Number of cores for parallel processing (default 1)
#' @return Data frame with taxon, effect (log2FC), pvalue, qvalue
run_deseq2 <- function(physeq, formula, contrast, nclusters = 1) {
  # Convert to DESeq2
  dds <- phyloseq_to_deseq2(physeq, as.formula(formula))

  # Set up parallel processing
  if (nclusters > 1) {
    BPPARAM <- BiocParallel::MulticoreParam(workers = nclusters)
  } else {
    BPPARAM <- BiocParallel::SerialParam()
  }

  # Run DESeq2 with geometric mean estimation that handles zeros
  dds <- DESeq2::estimateSizeFactors(dds, type = "poscounts")
  dds <- DESeq2::estimateDispersions(dds, fitType = "local")
  dds <- DESeq2::nbinomWaldTest(dds)

  # Get results
  res <- DESeq2::results(dds, contrast = contrast, parallel = nclusters > 1, BPPARAM = BPPARAM)

  data.frame(
    taxon = rownames(res),
    effect = res$log2FoldChange,
    pvalue = res$pvalue,
    qvalue = res$padj,
    method = "DESeq2",
    stringsAsFactors = FALSE
  ) |>
    filter(!is.na(effect))
}

#' Run edgeR on phyloseq object
#' @param physeq A phyloseq object
#' @param group_var Name of grouping variable in sample_data
#' @param contrast_levels Two-element vector: c(treatment, reference)
#' @return Data frame with taxon, effect (logFC), pvalue, qvalue
run_edgeR <- function(physeq, group_var, contrast_levels) {
  # Extract data
  counts <- as(otu_table(physeq), "matrix")
  if (!taxa_are_rows(physeq)) counts <- t(counts)

  group <- factor(sample_data(physeq)[[group_var]])
  group <- relevel(group, ref = contrast_levels[2])

  # Create DGEList
  dge <- edgeR::DGEList(counts = counts, group = group)

  # Filter low counts
  keep <- edgeR::filterByExpr(dge)
  dge <- dge[keep, , keep.lib.sizes = FALSE]

  # Normalize
  dge <- edgeR::calcNormFactors(dge, method = "TMM")

  # Design matrix
  design <- model.matrix(~ group)

 # Estimate dispersion
  dge <- edgeR::estimateDisp(dge, design)

  # Fit model
  fit <- edgeR::glmQLFit(dge, design)
  qlf <- edgeR::glmQLFTest(fit, coef = 2)

  # Get results
  res <- edgeR::topTags(qlf, n = Inf)$table

  data.frame(
    taxon = rownames(res),
    effect = res$logFC,
    pvalue = res$PValue,
    qvalue = res$FDR,
    method = "edgeR",
    stringsAsFactors = FALSE
  )
}

#' Run limma-voom on phyloseq object
#' @param physeq A phyloseq object
#' @param group_var Name of grouping variable in sample_data
#' @param contrast_levels Two-element vector: c(treatment, reference)
#' @return Data frame with taxon, effect (logFC), pvalue, qvalue
run_limma_voom <- function(physeq, group_var, contrast_levels) {
  # Extract data
  counts <- as(otu_table(physeq), "matrix")
  if (!taxa_are_rows(physeq)) counts <- t(counts)

  group <- factor(sample_data(physeq)[[group_var]])
  group <- relevel(group, ref = contrast_levels[2])

  # Create DGEList for normalization
  dge <- edgeR::DGEList(counts = counts, group = group)

  # Filter low counts
  keep <- edgeR::filterByExpr(dge)
  dge <- dge[keep, , keep.lib.sizes = FALSE]

  # Normalize
  dge <- edgeR::calcNormFactors(dge, method = "TMM")

  # Design matrix
  design <- model.matrix(~ group)

  # Voom transformation
  v <- limma::voom(dge, design, plot = FALSE)

  # Fit linear model
  fit <- limma::lmFit(v, design)
  fit <- limma::eBayes(fit)

  # Get results
  res <- limma::topTable(fit, coef = 2, number = Inf, sort.by = "none")

  data.frame(
    taxon = rownames(res),
    effect = res$logFC,
    pvalue = res$P.Value,
    qvalue = res$adj.P.Val,
    method = "limma-voom",
    stringsAsFactors = FALSE
  )
}

#' Run ALDEx2 on phyloseq object
#' @param physeq A phyloseq object
#' @param group_var Name of grouping variable in sample_data
#' @param contrast_levels Two-element vector: c(treatment, reference)
#' @param nclusters Number of cores for parallel processing (default 1)
#' @return Data frame with taxon, effect, pvalue, qvalue
run_aldex2 <- function(physeq, group_var, contrast_levels, nclusters = 1) {

  # Convert factor to character to avoid ALDEx2 error
  sample_data(physeq)[[group_var]] <- as.character(sample_data(physeq)[[group_var]])

  # Extract counts and conditions for direct ALDEx2 call
  counts <- as(otu_table(physeq), "matrix")
  if (!taxa_are_rows(physeq)) counts <- t(counts)

  conds <- sample_data(physeq)[[group_var]]

  # Run ALDEx2 with parallel support
  clr <- ALDEx2::aldex.clr(
    counts,
    conds,
    mc.samples = 128,
    useMC = nclusters > 1
  )
  tt <- ALDEx2::aldex.ttest(clr)
  effect <- ALDEx2::aldex.effect(clr, useMC = nclusters > 1)
  res <- cbind(tt, effect)

  data.frame(
    taxon = rownames(res),
    effect = res$effect,
    pvalue = res$wi.ep,
    qvalue = res$wi.eBH,
    method = "ALDEx2",
    stringsAsFactors = FALSE
  )
}

#' Run ANCOM-BC2 on phyloseq object
#' @param physeq A phyloseq object
#' @param group_var Name of grouping variable in sample_data
#' @param contrast_levels Two-element vector: c(treatment, reference)
#' @param nclusters Number of cores for parallel processing (default 1)
#' @return Data frame with taxon, effect (lfc), pvalue, qvalue
run_ancombc <- function(physeq, group_var, contrast_levels, nclusters = 1) {
  res <- MiscMetabar::ancombc_pq(
    physeq,
    fact = group_var,
    levels_fact = contrast_levels,
    tax_level = NULL,
    n_cl = nclusters
  )

  # Find the correct column names (depends on factor levels)
  lfc_col <- grep("^lfc_", colnames(res$res), value = TRUE)[1]
  p_col <- grep("^p_", colnames(res$res), value = TRUE)[1]
  q_col <- grep("^q_", colnames(res$res), value = TRUE)[1]

  data.frame(
    taxon = res$res$taxon,
    effect = res$res[[lfc_col]],
    pvalue = res$res[[p_col]],
    qvalue = res$res[[q_col]],
    method = "ANCOM-BC2",
    stringsAsFactors = FALSE
  ) |>
    filter(!is.na(effect))
}

#' Run MaAsLin3 on phyloseq object
#' @param physeq A phyloseq object
#' @param group_var Name of grouping variable in sample_data
#' @param contrast_levels Two-element vector: c(treatment, reference)
#' @param nclusters Number of cores for parallel processing (default 1)
#' @return Data frame with taxon, effect (coef), pvalue, qvalue
run_maaslin3 <- function(physeq, group_var, contrast_levels, nclusters = 1) {
  res <- maaslin3_pq(
    physeq,
    formula = paste0("~ ", group_var),
    reference = setNames(list(contrast_levels[2]), group_var),
    output = tempfile(),
    correction_for_sample_size = TRUE,
    plot_summary_plot = FALSE,
    plot_associations = FALSE,
    cores = nclusters
  )

  df <- res$fit_data_abundance$results |>
    filter(grepl(group_var, metadata))

  data.frame(
    taxon = df$feature,
    effect = df$coef,
    pvalue = df$pval_individual,
    qvalue = df$qval_individual,
    method = "MaAsLin3",
    stringsAsFactors = FALSE
  ) |>
    filter(!is.na(effect))
}

#' Run radEmu on phyloseq object
#' @param physeq A phyloseq object
#' @param group_var Name of grouping variable in sample_data
#' @param contrast_levels Two-element vector: c(treatment, reference)
#' @param return_wald_p If TRUE (default), use fast Wald p-values. If FALSE,
#'   use slower but more robust score tests with two-step filtering.
#' @param effect_threshold Minimum absolute effect to test with score tests
#'   (only used when return_wald_p = FALSE, default 0.5)
#' @param max_taxa_to_test Maximum number of taxa to test with score tests
#'   (only used when return_wald_p = FALSE, default NULL = all)
#' @param nclusters Number of cores for parallel processing (default 1)
#' @param ... Additional arguments passed to `radEmu::emuFit()`
#' @return Data frame with taxon, effect (estimate), pvalue, qvalue
run_emuFit <- function(physeq, 
group_var, 
contrast_levels,
                       return_wald_p = TRUE,
                       effect_threshold = 0.5,
                        max_taxa_to_test = NULL,
                       nclusters = 1, ...) {
  # Remove samples with zero total counts (radEmu requirement)
  sample_totals <- sample_sums(physeq)
  physeq <- prune_samples(sample_totals > 0, physeq)

  # Relevel factor so reference is first
  sample_data(physeq)[[group_var]] <- factor(
    sample_data(physeq)[[group_var]],
    levels = c(contrast_levels[2], contrast_levels[1])
  )

  if (return_wald_p) {
    # Fast approach: single fit with Wald p-values
    fit <- radEmu::emuFit(
      formula = as.formula(paste0("~ ", group_var)),
      Y = physeq,
      run_score_tests = FALSE,
      return_wald_p=return_wald_p,
      ...
    )

    # Extract results (Wald p-values are in pval column)
    coef_df <- fit$coef |>
      filter(covariate != "(Intercept)")

    # Adjust p-values for multiple testing
    coef_df$wald_q <- p.adjust(coef_df$wald_p, method = "BH")

    return(data.frame(
      taxon = coef_df$category,
      effect = coef_df$estimate,
      pvalue = coef_df$wald_p,
      qvalue = coef_df$wald_q,
      method = "radEmu_wald",
      stringsAsFactors = FALSE
    ) |>
      filter(!is.na(effect)))
  }

 # Two-step approach with score tests (slower but more robust)

  # Step 1: Run emuFit WITHOUT testing (fast) to get effect estimates
  fit_effects <- radEmu::emuFit(
    formula = as.formula(paste0("~ ", group_var)),
    Y = physeq,
    run_score_tests = FALSE,
    ...
  )

  # Extract effect estimates
  coef_df <- fit_effects$coef
  coef_df$abs_effect <- abs(coef_df$estimate)

  # Step 2: Select taxa with high effects to test
  # Filter by threshold and optionally limit to max_taxa_to_test
  taxa_to_test <- coef_df |>
    filter(covariate != "(Intercept)", abs_effect >= effect_threshold) |>
    arrange(desc(abs_effect))

  if (!is.null(max_taxa_to_test) && max_taxa_to_test > 0) {
    taxa_to_test <- head(taxa_to_test, max_taxa_to_test)
  }

  if (nrow(taxa_to_test) == 0) {
    # No taxa meet threshold - return all with NA p-values
    return(data.frame(
      taxon = coef_df$category[coef_df$covariate != "(Intercept)"],
      effect = coef_df$estimate[coef_df$covariate != "(Intercept)"],
      pvalue = NA_real_,
      qvalue = NA_real_,
      method = "radEmu",
      stringsAsFactors = FALSE
    ))
  }

  # Get indices of taxa to test (j values)
  taxa_names_vec <- taxa_names(physeq)
  j_to_test <- match(taxa_to_test$category, taxa_names_vec)

  # Step 3: Run emuFit WITH score testing only for selected taxa
  fit_tested <- radEmu::emuFit(
    formula = as.formula(paste0("~ ", group_var)),
    Y = physeq,
    test_kj = data.frame(k = 2, j = j_to_test),
    parallel = nclusters > 1,
    ncores = nclusters
  )

  # Extract tested results
  tested_res <- fit_tested$coef |>
    filter(covariate != "(Intercept)")

  # Merge with all effects
  all_effects <- coef_df |>
    filter(covariate != "(Intercept)") |>
    dplyr::select(category, estimate)

  res <- all_effects |>
    left_join(
      tested_res |> dplyr::select(category, pval),
      by = "category"
    )

  # Adjust p-values for multiple testing (only for tested taxa)
  res$qvalue <- p.adjust(res$pval, method = "BH")

  data.frame(
    taxon = res$category,
    effect = res$estimate,
    pvalue = res$pval,
    qvalue = res$qvalue,
    method = "radEmu_score",
    stringsAsFactors = FALSE
  ) |>
    filter(!is.na(effect))
}

#' Run all DA methods on a phyloseq object
#' @param physeq A phyloseq object
#' @param group_var Name of grouping variable
#' @param contrast_levels Two-element vector: c(treatment, reference)
#' @param methods Character vector of methods to run
#' @param effect_threshold Minimum effect for radEmu testing (default 2)
#' @param nclusters Number of cores for parallel processing (default 2)
#' @param verbose If TRUE, print method name and elapsed time (default FALSE)
#' @param max_taxa_to_test Maximum number of taxa to test for radEmu (default NULL)
#'   see `run_emuFit()`
#' @return Combined data frame of results
run_all_methods <- function(physeq, group_var, contrast_levels,
                            methods = c("DESeq2", "edgeR", "limma-voom",
                                       "ALDEx2", "ANCOM-BC2", "MaAsLin3", "radEmu"),
                            effect_threshold = 2,
                            nclusters = 2,
                            verbose = TRUE, 
                            max_taxa_to_test=0,
                            return_wald_p = TRUE) {
  results <- list()
  timings <- list()

  for (method in methods) {
    if (verbose) {
       cli::cli_alert(cli::col_red("Running", method, "..."))
    }

    start_time <- Sys.time()

    res <- tryCatch({
      switch(method,
        "DESeq2" = run_deseq2(physeq,
                             paste0("~ ", group_var),
                             c(group_var, contrast_levels[1], contrast_levels[2]),
                             nclusters = nclusters),
        "edgeR" = run_edgeR(physeq, group_var, contrast_levels),
        "limma-voom" = run_limma_voom(physeq, group_var, contrast_levels),
        "ALDEx2" = run_aldex2(physeq, group_var, contrast_levels,
                             nclusters = nclusters),
        "ANCOM-BC2" = run_ancombc(physeq, group_var, contrast_levels,
                                  nclusters = nclusters),
        "MaAsLin3" = run_maaslin3(physeq, group_var, contrast_levels,
                                  nclusters = nclusters),
        "radEmu" = run_emuFit(physeq, group_var, contrast_levels,
                              return_wald_p = return_wald_p,
                              effect_threshold = effect_threshold,
                              nclusters = nclusters,
                              max_taxa_to_test=max_taxa_to_test                              )
      )
    }, error = function(e) {
      warning("Error in ", method, ": ", e$message)
      NULL
    })

    elapsed <- as.numeric(difftime(Sys.time(), start_time, units = "secs"))
    timings[[method]] <- elapsed

    if (verbose) {
      cli::cli_alert(cli::col_red(" done (", round(elapsed, 1), "s)\n", sep = ""))
    }

    if (!is.null(res)) {
      results[[method]] <- res
    }
  }

  if (verbose) {
    cat("\nTotal time:", round(sum(unlist(timings)), 1), "s\n")
  }

  bind_rows(results)
}
```

## Dataset: Simulated Spike-in Data

To properly evaluate FDR control and sensitivity, we create datasets with
known differentially abundant taxa. There are three approaches available in
comparpq, each with different characteristics:

| Approach | Function | Key Feature |
|----------|----------|-------------|
| **Multiply with compensation** | `multiply_counts_pq()` | Multiplies counts then scales down non-selected taxa to preserve library size |
| **Permutation-based** | `permute_da_pq()` | Strictly preserves library sizes by redistributing counts |
| **MIDASim simulation** | `midasim_pq()` | Generates realistic simulated data preserving taxon correlations |

### Approach A: `multiply_counts_pq` with compensation

The `multiply_counts_pq()` function with `compensate = TRUE` creates a
compositional shift that DA methods can detect. Without compensation, the
multiplied samples would have larger library sizes, and DA methods would
normalize this away, washing out the signal.

```{r create_spike_data_mult}
data("data_fungi", package = "MiscMetabar")

# Subset to binary comparison
data_fungi_hl <- subset_samples(data_fungi, Height %in% c("Low", "High"))
data_fungi_hl <- prune_taxa(taxa_sums(data_fungi_hl) > 10, data_fungi_hl)

# Create spike-in with compensation to preserve library size
data_spike_mult <- multiply_counts_pq(
  data_fungi_hl,
  fact = "Height",
  conditions = "High",
  multipliers = 4,
  prop_taxa = 0.2,
  seed = 42,
  compensate = TRUE,
  min_prevalence = 0.1,
  verbose = TRUE
)

# Retrieve the actual modified taxa from the function output
spike_taxa_mult <- attr(data_spike_mult, "taxa_modified")

cat("Approach A - multiply_counts_pq with compensation:\n")
cat("- Total taxa:", ntaxa(data_spike_mult), "\n")
cat("- Spiked taxa:", length(spike_taxa_mult), "\n")
cat("- Samples:", nsamples(data_spike_mult), "\n")
cat("- Library size preserved:",
    all.equal(sample_sums(data_fungi_hl), sample_sums(data_spike_mult)), "\n")
```

### Approach B: `permute_da_pq` (strict library size preservation)

The `permute_da_pq()` function guarantees exact library size preservation
(except for rounding) by redistributing counts within each sample.

```{r create_spike_data_perm}
data_spike_perm <- permute_da_pq(
  data_fungi_hl,
  fact = "Height",
  conditions = "High",
  effect_size = 4,
  prop_taxa = 0.1,
  seed = 42,
  min_prevalence = 0.1
)

spike_taxa_perm <- attr(data_spike_perm, "da_taxa")

cat("\nApproach B - permute_da_pq:\n")
cat("- Spiked taxa:", length(spike_taxa_perm), "\n")
cat("- Library size preserved:",
    all.equal(sample_sums(data_fungi_hl), sample_sums(data_spike_perm)), "\n")
```

### Approach C: `midasim_pq` (realistic simulation)
```{r create_spike_data_mida, eval=requireNamespace("MIDASim", quietly = TRUE)}
data_spike_mida <- midasim_pq(
  data_fungi_hl,
  fact = "Height",
  condition = "High",
  effect_size = log(4),
  n_da_taxa = round(ntaxa(data_fungi_hl) * 0.1),
  seed = 42,
  min_prevalence = 0.1
)

spike_taxa_mida <- attr(data_spike_mida, "da_taxa_names")

cat("\nApproach C - midasim_pq:\n")
cat("- Spiked taxa:", length(spike_taxa_mida), "\n")
```

### Use Approach A for benchmark

For the main benchmark, we use the `multiply_counts_pq` approach with
compensation, as it modifies actual data while preserving library sizes.

```{r set_spike_data}
# Use the compensated multiply approach
data_spike <- clean_pq(data_spike_mult)
spike_taxa <- spike_taxa_mult

cat("\nDataset used for benchmark:\n")
cat("- Total taxa:", ntaxa(data_spike), "\n")
cat("- Spiked taxa:", length(spike_taxa), "\n")
cat("- Samples:", nsamples(data_spike), "\n")
cat("- Groups:", table(sample_data(data_spike)$Height), "\n")
```

## Run All Methods

```{r run_all_methods, results='hide'}
results_spike <- run_all_methods(
  subset_taxa_pq(data_spike, taxa_sums(data_spike) > 100),
  group_var = "Height",
  contrast_levels = c("High", "Low"),
  verbose=TRUE,
  ncluster=4
)
```

```{r show_results}
# Summary of results
results_spike |>
  group_by(method) |>
  summarise(
    n_taxa = n(),
    n_significant = sum(qvalue < 0.05, na.rm = TRUE),
    prop_significant = mean(qvalue < 0.05, na.rm = TRUE)
  ) |>
  knitr::kable(caption = "Summary of DA results by method", digits = 3)
```

## Concordance Analysis

### Number of Significant Taxa

```{r concordance_barplot, fig.height=5}
sig_summary <- results_spike |>
  mutate(significant = qvalue < 0.05) |>
  group_by(method) |>
  summarise(
    n_significant = sum(significant, na.rm = TRUE),
    n_tested = n()
  )

ggplot(sig_summary, aes(x = reorder(method, n_significant), y = n_significant)) +
  geom_col(fill = "steelblue") +
  geom_text(aes(label = n_significant), hjust = -0.2) +
  coord_flip() +
  labs(
    title = "Number of Significant Taxa by Method",
    subtitle = "q-value < 0.05",
    x = NULL,
    y = "Number of significant taxa"
  ) +
  theme_bw() +
  expand_limits(y = max(sig_summary$n_significant) * 1.1)
```

### Overlap Between Methods (UpSet Plot)

```{r upset_plot, fig.height=6, fig.width=10}
# Create presence/absence matrix for significant taxa
sig_taxa_list <- results_spike |>
  filter(qvalue < 0.05) |>
  group_by(method) |>
  summarise(taxa = list(taxon)) |>
  tibble::deframe()

# Convert to binary matrix
all_taxa <- unique(unlist(sig_taxa_list))
sig_matrix <- sapply(sig_taxa_list, function(x) all_taxa %in% x)
rownames(sig_matrix) <- all_taxa

if (nrow(sig_matrix) > 0) {
  # UpSet-style visualization
  overlap_df <- as.data.frame(sig_matrix) |>
    rownames_to_column("taxon") |>
    pivot_longer(-taxon, names_to = "method", values_to = "significant") |>
    filter(significant) |>
    group_by(taxon) |>
    summarise(
      n_methods = n(),
      methods = paste(sort(method), collapse = " + ")
    )

  ggplot(overlap_df, aes(x = reorder(methods, n_methods), fill = factor(n_methods))) +
    geom_bar() +
    coord_flip() +
    scale_fill_viridis_d(name = "# Methods") +
    labs(
      title = "Taxa Overlap Across Methods",
      x = "Method combination",
      y = "Number of taxa"
    ) +
    theme_bw() +
    theme(legend.position = "right")
}
```

### Pairwise Method Agreement

```{r pairwise_agreement, fig.height=7, fig.width=8}
# Calculate Jaccard similarity between methods
methods <- unique(results_spike$method)
jaccard_matrix <- matrix(NA, length(methods), length(methods),
                        dimnames = list(methods, methods))

for (i in seq_along(methods)) {
  for (j in seq_along(methods)) {
    taxa_i <- results_spike$taxon[results_spike$method == methods[i] &
                                   results_spike$qvalue < 0.05]
    taxa_j <- results_spike$taxon[results_spike$method == methods[j] &
                                   results_spike$qvalue < 0.05]

    intersection <- length(intersect(taxa_i, taxa_j))
    union <- length(union(taxa_i, taxa_j))

    jaccard_matrix[i, j] <- if (union > 0) intersection / union else 0
  }
}

# Plot heatmap
jaccard_df <- as.data.frame(as.table(jaccard_matrix))
names(jaccard_df) <- c("Method1", "Method2", "Jaccard")

ggplot(jaccard_df, aes(x = Method1, y = Method2, fill = Jaccard)) +
  geom_tile() +
  geom_text(aes(label = round(Jaccard, 2)), color = "white", size = 4) +
  scale_fill_viridis_c(limits = c(0, 1)) +
  labs(
    title = "Pairwise Method Agreement (Jaccard Similarity)",
    subtitle = "Based on significant taxa (q < 0.05)",
    x = NULL, y = NULL
  ) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Effect Size Comparison

### Effect Size Correlation

```{r effect_correlation, fig.height=8, fig.width=10}
# Wide format for effect sizes
effects_wide <- results_spike |>
  dplyr::select(taxon, method, effect) |>
  pivot_wider(names_from = method, values_from = effect)

# Pairwise scatter plots
if (ncol(effects_wide) > 2) {
  # Calculate correlations
  cor_matrix <- cor(effects_wide[, -1], use = "pairwise.complete.obs")

  # Plot correlation matrix
  cor_df <- as.data.frame(as.table(cor_matrix))
  names(cor_df) <- c("Method1", "Method2", "Correlation")

  p_cor <- ggplot(cor_df, aes(x = Method1, y = Method2, fill = Correlation)) +
    geom_tile() +
    geom_text(aes(label = round(Correlation, 2)), color = "black", size = 3) +
    scale_fill_gradient2(low = "blue", mid = "white", high = "red",
                        midpoint = 0, limits = c(-1, 1)) +
    labs(
      title = "Effect Size Correlation Between Methods",
      x = NULL, y = NULL
    ) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

  print(p_cor)
}
```

### Effect Size Distribution

```{r effect_distribution, fig.height=6}
ggplot(results_spike, aes(x = effect, fill = method)) +
  geom_density(alpha = 0.5) +
  facet_wrap(~method, scales = "free_y") +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(
    title = "Distribution of Effect Sizes by Method",
    x = "Effect Size (log2 fold change)",
    y = "Density"
  ) +
  theme_bw() +
  theme(legend.position = "none")
```

## FDR Control Assessment

Using our spike-in data, we can evaluate how well each method controls FDR.

```{r fdr_assessment}
# Get the actual spiked taxa (those with multiplied counts)
# We need to identify them from the multiply_counts_pq function
# For this, we'll use the taxa that were selected (stored during creation)

# Mark true positives
results_spike <- results_spike |>
  mutate(
    true_da = taxon %in% spike_taxa,
    called_da = qvalue < 0.05
  )

# Calculate performance metrics
performance <- results_spike |>
  group_by(method) |>
  summarise(
    TP = sum(true_da & called_da, na.rm = TRUE),
    FP = sum(!true_da & called_da, na.rm = TRUE),
    TN = sum(!true_da & !called_da, na.rm = TRUE),
    FN = sum(true_da & !called_da, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(
    Sensitivity = TP / (TP + FN),
    Specificity = TN / (TN + FP),
    FDR = FP / (TP + FP),
    Precision = TP / (TP + FP),
    F1 = 2 * (Precision * Sensitivity) / (Precision + Sensitivity)
  )

knitr::kable(
  performance |> 
    dplyr::select(method, TP, FP, FN, Sensitivity, FDR, F1),
  caption = "Performance metrics based on spike-in ground truth",
  digits = 3
)
```

```{r performance_plot, fig.height=5, fig.width=9}
performance_long <- performance |>
  dplyr::select(method, Sensitivity, Specificity, FDR, F1) |>
  pivot_longer(-method, names_to = "Metric", values_to = "Value")

ggplot(performance_long, aes(x = method, y = Value, fill = Metric)) +
  geom_col(position = "dodge") +
  geom_hline(data = data.frame(Metric = "FDR", y = 0.05),
             aes(yintercept = y), linetype = "dashed", color = "red") +
  facet_wrap(~Metric, scales = "free_y") +
  labs(
    title = "Performance Metrics by Method",
    subtitle = "Red dashed line = nominal FDR (0.05)",
    x = NULL, y = "Value"
  ) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none")
```

## Volcano Plots Comparison

```{r volcano_comparison, fig.height=10, fig.width=12}
results_spike <- results_spike |>
  mutate(
    neg_log10_q = -log10(qvalue),
    significant = qvalue < 0.05,
    category = case_when(
      true_da & significant ~ "True Positive",
      !true_da & significant ~ "False Positive",
      true_da & !significant ~ "False Negative",
      TRUE ~ "True Negative"
    )
  )

ggplot(results_spike, aes(x = effect, y = neg_log10_q, color = category)) +
  geom_point(alpha = 0.6, size = 1.5) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey40") +
  geom_vline(xintercept = 0, color = "grey40") +
  scale_color_manual(values = c(
    "True Positive" = "forestgreen",
    "False Positive" = "red",
    "False Negative" = "orange",
    "True Negative" = "grey70"
  )) +
  facet_wrap(~method, scales = "free") +
  labs(
    title = "Volcano Plots by Method",
    subtitle = "Colors based on spike-in ground truth",
    x = "Effect Size",
    y = "-log10(q-value)",
    color = "Classification"
  ) +
  theme_bw() +
  theme(legend.position = "bottom")
```

## Real Data: HMP2 Antibiotics

Now we apply the methods to real data without known ground truth.

```{r hmp2_data}
# Load HMP2 data
taxa_table_name <- system.file("extdata", "HMP2_taxonomy.tsv", package = "maaslin3")
taxa_table <- read.csv(taxa_table_name, sep = "\t", row.names = 1)

metadata_name <- system.file("extdata", "HMP2_metadata.tsv", package = "maaslin3")
metadata <- read.csv(metadata_name, sep = "\t", row.names = 1)

metadata$antibiotics <- factor(metadata$antibiotics, levels = c("No", "Yes"))

# Create phyloseq
otu <- otu_table(as.matrix(taxa_table), taxa_are_rows = FALSE)
sam <- sample_data(metadata)
species_names <- colnames(taxa_table)
tax_df <- data.frame(
  Species = species_names,
  Genus = sapply(strsplit(species_names, "_"), \(x) x[1]),
  row.names = species_names
)
tax <- tax_table(as.matrix(tax_df))
physeq_hmp2 <- phyloseq(otu, sam, tax)

# Convert to integer counts for count-based methods
# HMP2 data appears to be relative abundances, multiply by 1e6
otu_table(physeq_hmp2) <- otu_table(
  round(as.matrix(otu_table(physeq_hmp2)) * 1e6),
  taxa_are_rows = FALSE
)

cat("HMP2 Dataset:\n")
cat("- Taxa:", ntaxa(physeq_hmp2), "\n")
cat("- Samples:", nsamples(physeq_hmp2), "\n")
```

```{r run_hmp2, results='hide'}
results_hmp2 <- run_all_methods(
  physeq_hmp2,
  methods = c("ALDEx2", "ANCOM-BC2", "MaAsLin3", "radEmu"),
  group_var = "antibiotics",
  contrast_levels = c("Yes", "No"),
  verbose=TRUE,
  nclusters = 4
)
```

```{r hmp2_summary, fig.height=8, fig.width=10}
# Concordance plot
sig_hmp2 <- results_hmp2 |>
  filter(qvalue < 0.05) |>
  group_by(method) |>
  summarise(taxa = list(taxon)) |>
  tibble::deframe()

# Calculate consensus
consensus_hmp2 <- results_hmp2 |>
  filter(qvalue < 0.05) |>
  group_by(taxon) |>
  summarise(
    n_methods = n_distinct(method),
    methods = paste(sort(unique(method)), collapse = ", "),
    mean_effect = mean(effect, na.rm = TRUE)
  ) |>
  arrange(desc(n_methods), desc(abs(mean_effect)))

cat("\nTop consensus taxa (found by 4+ methods):\n")
consensus_hmp2 |>
  filter(n_methods >= 2) |>
  head(15) |>
  knitr::kable(digits = 2)
```

```{r hmp2_volcano, fig.height=8}
results_hmp2 |>
  mutate(significant = qvalue < 0.05) |>
  ggplot(aes(x = effect, y = -log10(qvalue), color = significant)) +
  geom_point(alpha = 0.5) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed") +
  scale_color_manual(values = c("FALSE" = "grey60", "TRUE" = "firebrick")) +
  facet_wrap(~method, scales = "free") +
  labs(
    title = "HMP2: Antibiotics Effect",
    x = "Effect Size",
    y = "-log10(q-value)"
  ) +
  theme_bw() +
  theme(legend.position = "bottom")
```

## Session Info

```{r session_info}
sessionInfo()
```
