---
title: "Comparing Differential Abundance Methods"
author: "Adrien Taudière"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comparing Differential Abundance Methods}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.width = 9,
  fig.height = 6
)
```

## Introduction

This vignette compares three popular differential abundance (DA) methods for
microbiome data:

- **ANCOM-BC2** (Analysis of Compositions of Microbiomes with Bias Correction):
  Uses a linear regression framework with bias correction for compositionality
- **ALDEx2** (ANOVA-Like Differential Expression): Uses centered log-ratio
  transformation and Monte Carlo sampling from Dirichlet distribution
- **MaAsLin3** (Microbiome Multivariable Associations with Linear Models):
  Fits generalized linear models with options for both abundance and prevalence

We compare these methods on three different datasets to illustrate their
behavior across different experimental designs and data characteristics.

## Setup

```{r setup}
library(comparpq)
library(phyloseq)
library(ggplot2)
library(dplyr)
library(patchwork)
```

## Dataset 1: Fungi and Tree Height

The `data_fungi` dataset from MiscMetabar contains fungal communities sampled
at different heights on trees. We compare Low vs High height positions.

### Data Preparation

```{r data_fungi_prep}
data("data_fungi", package = "MiscMetabar")

# Subset to Low and High only for a clear binary comparison
data_fungi_hl <- subset_samples(data_fungi, Height %in% c("Low", "High"))
data_fungi_hl <- prune_taxa(taxa_sums(data_fungi_hl) > 0, data_fungi_hl)

cat("Dataset: data_fungi (Low vs High)\n")
cat("Samples:", nsamples(data_fungi_hl), "\n")
cat("Taxa:", ntaxa(data_fungi_hl), "\n")
cat("Groups:", table(sample_data(data_fungi_hl)$Height), "\n")
```

### Run ANCOM-BC

```{r fungi_ancombc, results='hide'}
res_ancombc_fungi <- MiscMetabar::ancombc_pq(
  data_fungi_hl,
  fact = "Height",
  levels_fact = c("Low", "High"),
  tax_level = NULL
)
```

```{r fungi_ancombc_show}
# Extract results
ancombc_df_fungi <- res_ancombc_fungi$res |>
  filter(!is.na(lfc_HeightHigh)) |>
  mutate(
    method = "ANCOM-BC2",
    effect = lfc_HeightHigh,
    qvalue = q_HeightHigh,
    significant = diff_HeightHigh
  ) |>
  dplyr::select(taxon, method, effect, qvalue, significant)

cat("ANCOM-BC significant taxa:", sum(ancombc_df_fungi$significant), "\n")
```

### Run ALDEx2

```{r fungi_aldex, results='hide'}
res_aldex_fungi <- MiscMetabar::aldex_pq(
  data_fungi_hl,
  bifactor = "Height",
  modalities = c("Low", "High")
)
```

```{r fungi_aldex_show}
aldex_df_fungi <- res_aldex_fungi |>
  tibble::rownames_to_column("taxon") |>
  mutate(
    method = "ALDEx2",
    effect = effect,
    qvalue = wi.eBH,
    significant = wi.eBH < 0.05
  ) |>
  dplyr::select(taxon, method, effect, qvalue, significant)

cat("ALDEx2 significant taxa:", sum(aldex_df_fungi$significant), "\n")
```

### Run MaAsLin3

```{r fungi_maaslin3, results='hide'}
res_maaslin3_fungi <- maaslin3_pq(
  data_fungi_hl,
  formula = "~ Height",
  reference = list(Height = "Low"),
  output = tempfile(),
  correction_for_sample_size = FALSE,
  plot_summary_plot = FALSE,
  plot_associations = FALSE
)
```

```{r fungi_maaslin3_show}
maaslin3_df_fungi <- res_maaslin3_fungi$fit_data_abundance$results |>
  filter(metadata == "Height") |>
  mutate(
    taxon = feature,
    method = "MaAsLin3",
    effect = coef,
    qvalue = qval_individual,
    significant = qval_individual < 0.05
  ) |>
  dplyr::select(taxon, method, effect, qvalue, significant)

cat("MaAsLin3 significant taxa:", sum(maaslin3_df_fungi$significant, na.rm = TRUE), "\n")
```

### Compare Results

```{r fungi_compare, fig.height=8}
# Combine results
combined_fungi <- bind_rows(ancombc_df_fungi, aldex_df_fungi, maaslin3_df_fungi)

# Volcano plots
p_volcano_fungi <- ggplot(
  combined_fungi,
  aes(x = effect, y = -log10(qvalue), color = significant)
) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey40") +
  geom_vline(xintercept = 0, color = "grey40") +
  scale_color_manual(values = c("FALSE" = "grey60", "TRUE" = "firebrick3")) +
  facet_wrap(~method, scales = "free") +
  labs(
    title = "Data Fungi: Low vs High Height",
    x = "Effect Size",
    y = "-log10(q-value)"
  ) +
  theme_bw() +
  theme(legend.position = "bottom") +
  geom_label(
    data = combined_fungi %>% filter(significant),
    aes(label=taxon)
  )

print(p_volcano_fungi)
```

```{r fungi_venn}
# Find overlapping significant taxa
sig_ancombc <- ancombc_df_fungi$taxon[ancombc_df_fungi$significant]
sig_aldex <- aldex_df_fungi$taxon[aldex_df_fungi$significant]
sig_maaslin3 <- na.omit(maaslin3_df_fungi$taxon[maaslin3_df_fungi$significant])

# Overlap summary
cat("\nSignificant taxa overlap:\n")
cat("ANCOM-BC only:", length(setdiff(sig_ancombc, union(sig_aldex, sig_maaslin3))), "\n")
cat("ALDEx2 only:", length(setdiff(sig_aldex, union(sig_ancombc, sig_maaslin3))), "\n")
cat("MaAsLin3 only:", length(setdiff(sig_maaslin3, union(sig_ancombc, sig_aldex))), "\n")
cat("All three:", length(Reduce(intersect, list(sig_ancombc, sig_aldex, sig_maaslin3))), "\n")
```

## Dataset 2: HMP2 and Antibiotics

The HMP2 dataset from the maaslin3 package contains gut microbiome data from
the Human Microbiome Project 2. We compare samples with and without antibiotic
usage.

### Data Preparation

```{r hmp2_prep}
# Read the HMP2 default dataset from maaslin3 package
taxa_table_name <- system.file("extdata", "HMP2_taxonomy.tsv", package = "maaslin3")
taxa_table <- read.csv(taxa_table_name, sep = "\t", row.names = 1)

metadata_name <- system.file("extdata", "HMP2_metadata.tsv", package = "maaslin3")
metadata <- read.csv(metadata_name, sep = "\t", row.names = 1)

# Set factor levels
metadata$antibiotics <- factor(metadata$antibiotics, levels = c("No", "Yes"))

# Create phyloseq object
otu <- otu_table(as.matrix(taxa_table), taxa_are_rows = FALSE)
sam <- sample_data(metadata)
species_names <- colnames(taxa_table)
tax_df <- data.frame(
  Species = species_names,
  Genus = sapply(strsplit(species_names, "_"), \(x) x[1]),
  row.names = species_names
)
tax <- tax_table(as.matrix(tax_df))
physeq_hmp2 <- phyloseq(otu, sam, tax)

cat("Dataset: HMP2 (Antibiotics No vs Yes)\n")
cat("Samples:", nsamples(physeq_hmp2), "\n")
cat("Taxa:", ntaxa(physeq_hmp2), "\n")
cat("Groups:", table(sample_data(physeq_hmp2)$antibiotics), "\n")
```

### Run ANCOM-BC

```{r hmp2_ancombc, results='hide'}
res_ancombc_hmp2 <- MiscMetabar::ancombc_pq(
  physeq_hmp2,
  fact = "antibiotics",
  levels_fact = c("No", "Yes"),
  tax_level = NULL
)
```

```{r hmp2_ancombc_show}
ancombc_df_hmp2 <- res_ancombc_hmp2$res |>
  filter(!is.na(lfc_antibioticsYes)) |>
  mutate(
    method = "ANCOM-BC",
    effect = lfc_antibioticsYes,
    qvalue = q_antibioticsYes,
    significant = diff_antibioticsYes
  ) |>
  dplyr::select(taxon, method, effect, qvalue, significant)

cat("ANCOM-BC significant taxa:", sum(ancombc_df_hmp2$significant), "\n")
```

### Run ALDEx2

```{r hmp2_aldex, results='hide'}
physeq_hmp2@sam_data$antibiotics <- as.character(physeq_hmp2@sam_data$antibiotics)
physeq_hmp2@otu_table <- otu_table(
  round(as.matrix(physeq_hmp2@otu_table))*1000,
  taxa_are_rows = taxa_are_rows(physeq_hmp2)
)

res_aldex_hmp2 <- MiscMetabar::aldex_pq(
  physeq_hmp2,
  test="t",
  bifactor = "antibiotics",
  modalities = c("No", "Yes")
)
```

```{r hmp2_aldex_show}
aldex_df_hmp2 <- res_aldex_hmp2 |>
  tibble::rownames_to_column("taxon") |>
  mutate(
    method = "ALDEx2",
    effect = effect,
    qvalue = wi.eBH,
    significant = wi.eBH < 0.05
  ) |>
  dplyr::select(taxon, method, effect, qvalue, significant)

cat("ALDEx2 significant taxa:", sum(aldex_df_hmp2$significant), "\n")
```

### Run MaAsLin3

```{r hmp2_maaslin3, results='hide'}
otu_maaslin <- otu_table(as.matrix(taxa_table), taxa_are_rows = FALSE)
sam_maaslin <- sample_data(metadata)
physeq_hmp2_maaslin <- phyloseq(otu_maaslin, sam_maaslin, tax)

res_maaslin3_hmp2 <- maaslin3_pq(
  physeq_hmp2_maaslin,
  formula = "~ antibiotics",
  reference = list(antibiotics = "No"),
  output = tempfile()
)
```

```{r hmp2_maaslin3_show}
maaslin3_df_hmp2 <- res_maaslin3_hmp2$fit_data_abundance$results |>
  filter(metadata == "antibiotics") |>
  mutate(
    taxon = feature,
    method = "MaAsLin3",
    effect = coef,
    qvalue = qval_individual,
    significant = qval_individual < 0.05
  ) |>
  dplyr::select(taxon, method, effect, qvalue, significant)

cat("MaAsLin3 significant taxa:", sum(maaslin3_df_hmp2$significant, na.rm = TRUE), "\n")
```

### Compare Results

```{r hmp2_compare, fig.height=8}
combined_hmp2 <- bind_rows(ancombc_df_hmp2, aldex_df_hmp2, maaslin3_df_hmp2)

p_volcano_hmp2 <- ggplot(
  combined_hmp2,
  aes(x = effect, y = -log10(qvalue), color = significant)
) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey40") +
  geom_vline(xintercept = 0, color = "grey40") +
  scale_color_manual(values = c("FALSE" = "grey60", "TRUE" = "firebrick3")) +
  facet_wrap(~method, scales = "free") +
  labs(
    title = "HMP2: Antibiotics No vs Yes",
    x = "Effect Size",
    y = "-log10(q-value)"
  ) +
  theme_bw() +
  theme(legend.position = "bottom") +
  geom_label(
    data = combined_hmp2 %>% filter(significant),
    aes(label=taxon)
  )

print(p_volcano_hmp2)
```

### MaAsLin3 Summary Plot

```{r hmp2_summary, fig.width=10, fig.height=8}
# Use the default summary plot from gg_maaslin3_plot
p_summary <- gg_maaslin3_plot(res_maaslin3_hmp2, type = "summary", top_n = 20)
print(p_summary)
```

## Dataset 3: HMP Gingival Microbiome

The HMP_2012_16S_gingival_V13 dataset from MicrobiomeBenchmarkData contains
oral microbiome data comparing subgingival and supragingival plaque.

### Data Preparation

```{r gingival_prep}
# Get the dataset
gingival_data <- MicrobiomeBenchmarkData::getBenchmarkData(
  "HMP_2012_16S_gingival_V13",
  dryrun = FALSE
)[[1]]

# Convert to phyloseq
gingival_pq <- mia::convertToPhyloseq(gingival_data)
gingival_pq@tax_table <- tax_table(data.frame(
  Genus = paste0("Genus", 1:ntaxa(gingival_pq)),
  row.names = taxa_names(gingival_pq)
))

taxa_names(gingival_pq) <- names(gingival_data)

# Subset gingival_pq to only 1000 taxa for efficiency
set.seed(123)
selected_taxa <- sample(taxa_names(gingival_pq), 1000)
gingival_pq <- prune_taxa(selected_taxa, gingival_pq)

cat("Dataset: HMP Gingival (subgingival vs supragingival plaque)\n")
cat("Samples:", nsamples(gingival_pq), "\n")
cat("Taxa:", ntaxa(gingival_pq), "\n")
cat("Groups:", table(sample_data(gingival_pq)$body_subsite), "\n")
```

### Run ANCOM-BC

```{r gingival_ancombc, results='hide'}
res_ancombc_ging <- MiscMetabar::ancombc_pq(
  gingival_pq,
  fact = "body_subsite",
  levels_fact = c("subgingival_plaque", "supragingival_plaque"),
  tax_level = NULL
)
```

```{r gingival_ancombc_show}
# Column name varies based on factor level
lfc_col <- grep("^lfc_", colnames(res_ancombc_ging$res), value = TRUE)[1]
q_col <- grep("^q_", colnames(res_ancombc_ging$res), value = TRUE)[1]
diff_col <- grep("^diff_", colnames(res_ancombc_ging$res), value = TRUE)[1]

ancombc_df_ging <- res_ancombc_ging$res |>
  filter(!is.na(.data[[lfc_col]])) |>
  mutate(
    method = "ANCOM-BC",
    effect = .data[[lfc_col]],
    qvalue = .data[[q_col]],
    significant = .data[[diff_col]]
  ) |>
  dplyr::select(taxon, method, effect, qvalue, significant)

cat("ANCOM-BC significant taxa:", sum(ancombc_df_ging$significant), "\n")
```

### Run ALDEx2

```{r gingival_aldex, results='hide'}
res_aldex_ging <- MiscMetabar::aldex_pq(
  gingival_pq,
  bifactor = "body_subsite",
  modalities = c("subgingival_plaque", "supragingival_plaque")
)
```

```{r gingival_aldex_show}
aldex_df_ging <- res_aldex_ging |>
  tibble::rownames_to_column("taxon") |>
  mutate(
    method = "ALDEx2",
    effect = effect,
    qvalue = wi.eBH,
    significant = wi.eBH < 0.05
  ) |>
  dplyr::select(taxon, method, effect, qvalue, significant)

cat("ALDEx2 significant taxa:", sum(aldex_df_ging$significant), "\n")
```

### Run MaAsLin3

```{r gingival_maaslin3, results='hide'}
res_maaslin3_ging <- maaslin3_pq(
  gingival_pq,
  formula = "~ body_subsite",
  reference = list(body_subsite = "subgingival_plaque"),
  output = tempfile(),
  correction_for_sample_size = FALSE,
  plot_summary_plot = FALSE,
  plot_associations = FALSE
)
```

```{r gingival_maaslin3_show}
maaslin3_df_ging <- res_maaslin3_ging$fit_data_abundance$results |>
  filter(grepl("body_subsite", metadata)) |>
  mutate(
    taxon = feature,
    method = "MaAsLin3",
    effect = coef,
    qvalue = qval_individual,
    significant = qval_individual < 0.05
  ) |>
  dplyr::select(taxon, method, effect, qvalue, significant)

cat("MaAsLin3 significant taxa:", sum(maaslin3_df_ging$significant, na.rm = TRUE), "\n")
```

### Compare Results

```{r gingival_compare, fig.height=8}
combined_ging <- bind_rows(ancombc_df_ging, aldex_df_ging, maaslin3_df_ging)

p_volcano_ging <- ggplot(
  combined_ging,
  aes(x = effect, y = -log10(qvalue), color = significant)
) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey40") +
  geom_vline(xintercept = 0, color = "grey40") +
  scale_color_manual(values = c("FALSE" = "grey60", "TRUE" = "firebrick3")) +
  facet_wrap(~method, scales = "free") +
  labs(
    title = "HMP Gingival: Subgingival vs Supragingival Plaque",
    x = "Effect Size",
    y = "-log10(q-value)"
  ) +
  theme_bw() +
  theme(legend.position = "bottom") +
  geom_label(
    data = combined_ging %>% filter(qvalue < 0.0001),
    aes(label=taxon)
  )

print(p_volcano_ging)
```



## Dataset 4: selex

```{r selex_prep}
library(ALDEx2)
data(selex)
selex <- selex[1201:1300,] # subset for efficiency
conds <- c(rep("NS", 7), rep("S", 7))

physeq_selex <- phyloseq(otu_table(selex, taxa_are_rows = TRUE), 
          sample_data(data.frame(condition = conds, row.names = colnames(selex))))

physeq_selex@tax_table <-  tax_table(data.frame(Genus = paste0("Genus", 1:100), row.names = rownames(selex)))

taxa_names(physeq_selex) <- rownames(selex)
```

### Run ANCOM-BC

```{r selex_ancombc, results='hide'}
res_ancombc_selex <- MiscMetabar::ancombc_pq(
  physeq_selex,
  fact = "condition",
  levels_fact = c("S", "NS"),
  tax_level = NULL
)
```

```{r selex_ancombc_show}
ancombc_df_selex <- res_ancombc_selex$res |>
  filter(!is.na(lfc_conditionNS)) |>
  mutate(
    method = "ANCOM-BC",
    effect = lfc_conditionNS,
    qvalue = q_conditionNS,
    significant = diff_conditionNS
  ) |>
  dplyr::select(taxon, method, effect, qvalue, significant)

cat("ANCOM-BC significant taxa:", sum(ancombc_df_selex$significant), "\n")
```

### Run ALDEx2

```{r selex_aldex, results='hide'}
res_aldex_selex <- MiscMetabar::aldex_pq(
  physeq_selex,
  test="t",
  bifactor = "condition",
  modalities = c("S", "NS")
)
```

```{r selex_aldex_show}
aldex_df_selex <- res_aldex_selex |>
  tibble::rownames_to_column("taxon") |>
  mutate(
    method = "ALDEx2",
    effect = effect,
    qvalue = wi.eBH,
    significant = wi.eBH < 0.05
  ) |>
  dplyr::select(taxon, method, effect, qvalue, significant)

cat("ALDEx2 significant taxa:", sum(aldex_df_selex$significant), "\n")
```

### Run MaAsLin3

```{r selex_maaslin3, results='hide'}
res_maaslin3_selex <- maaslin3_pq(
  physeq_selex,
  formula = "~ condition",
  reference = list(S = "S"),
  output = tempfile()
)
```

```{r selex_maaslin3_show}
maaslin3_df_selex <- res_maaslin3_selex$fit_data_abundance$results |>
  filter(metadata == "condition") |>
  mutate(
    taxon = feature,
    method = "MaAsLin3",
    effect = coef,
    qvalue = qval_individual,
    significant = qval_individual < 0.05
  ) |>
  dplyr::select(taxon, method, effect, qvalue, significant)

cat("MaAsLin3 significant taxa:", sum(maaslin3_df_selex$significant, na.rm = TRUE), "\n")
```

### Compare Results

```{r selex_compare, fig.height=8}
combined_selex <- bind_rows(ancombc_df_selex, aldex_df_selex, maaslin3_df_selex)

p_volcano_selex <- ggplot(
  combined_selex,
  aes(x = effect, y = -log10(qvalue), color = significant)
) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey40") +
  geom_vline(xintercept = 0, color = "grey40") +
  scale_color_manual(values = c("FALSE" = "grey60", "TRUE" = "firebrick3")) +
  facet_wrap(~method, scales = "free") +
  labs(
    title = "selex: condition S vs NS",
    x = "Effect Size",
    y = "-log10(q-value)"
  ) +
  theme_bw() +
  theme(legend.position = "bottom") +
  geom_label(
    data = combined_selex %>% filter(significant),
    aes(label=taxon)
  )

print(p_volcano_selex)
```

```{r}
combined_selex |>
  filter(significant) |>
  group_by(taxon) |> 
  summarise(method_signif=n(), method=paste(method, collapse="; ")) |>
  arrange(method_signif)
```





## Summary Across Datasets

```{r summary_table}
# Create summary table
summary_df <- tibble::tribble(
  ~Dataset, ~Method, ~`Total Taxa`, ~`Significant Taxa`,
  "Fungi (Height)", "ANCOM-BC", nrow(ancombc_df_fungi), sum(ancombc_df_fungi$significant),
  "Fungi (Height)", "ALDEx2", nrow(aldex_df_fungi), sum(aldex_df_fungi$significant),
  "Fungi (Height)", "MaAsLin3", nrow(maaslin3_df_fungi), sum(maaslin3_df_fungi$significant, na.rm = TRUE),
  "HMP2 (Antibiotics)", "ANCOM-BC", nrow(ancombc_df_hmp2), sum(ancombc_df_hmp2$significant),
  "HMP2 (Antibiotics)", "ALDEx2", nrow(aldex_df_hmp2), sum(aldex_df_hmp2$significant),
  "HMP2 (Antibiotics)", "MaAsLin3", nrow(maaslin3_df_hmp2), sum(maaslin3_df_hmp2$significant, na.rm = TRUE),
  "Gingival (Body Site)", "ANCOM-BC", nrow(ancombc_df_ging), sum(ancombc_df_ging$significant),
  "Gingival (Body Site)", "ALDEx2", nrow(aldex_df_ging), sum(aldex_df_ging$significant),
  "Gingival (Body Site)", "MaAsLin3", nrow(maaslin3_df_ging), sum(maaslin3_df_ging$significant, na.rm = TRUE)
)

knitr::kable(summary_df, caption = "Summary of Differential Abundance Results")
```

```{r summary_plot, fig.height=5, fig.width=8}
summary_df |>
  mutate(Proportion = `Significant Taxa` / `Total Taxa`) |>
  ggplot(aes(x = Method, y = `Significant Taxa`, fill = Dataset)) +
  geom_col(position = "dodge") +
  labs(
    title = "Number of Significant Taxa by Method and Dataset",
    y = "Number of Significant Taxa"
  ) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Discussion

The three methods show different sensitivities across datasets:

1. **ANCOM-BC** tends to detect more significant taxa, especially in datasets
   with clear biological differences. It accounts for sampling fraction bias
   and provides log fold change estimates.

2. **ALDEx2** is generally more conservative, using Monte Carlo sampling from
   the Dirichlet distribution to account for uncertainty. It tends to have
   fewer false positives but may miss some true effects.

3. **MaAsLin3** provides a flexible framework with both abundance and
   prevalence models. Its sensitivity depends on the transformation and
   normalization settings used.

Key considerations when choosing a method:

- **Sample size**: ALDEx2 may be less powerful with small sample sizes
- **Compositionality**: All three methods address this, but differently
- **Effect size interpretation**: ANCOM-BC and MaAsLin3 provide log fold changes
- **Computational time**: ALDEx2 can be slower due to Monte Carlo sampling

For robust results, consider using multiple methods and focusing on taxa
identified by at least two methods.

## Session Info

```{r session}
sessionInfo()
```
