---
title: "Estimation Statistics for Microbiome Data"
author: "Adrien Taudière"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Estimation Statistics for Microbiome Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.width = 9,
  fig.height = 6
)
```

## Why estimation statistics?

Null hypothesis significance testing (NHST) has been the dominant paradigm in
biological research for decades, yet its limitations have been discussed for
over 77 years. The core issue is that a p-value answers a narrow question ---
"Is the effect non-zero?" --- while telling us nothing about **how large**
the effect is or **how precise** our estimate is. As
[Ho et al. (2019)](https://www.nature.com/articles/s41592-019-0470-3) put it:

> Although NHST limits the analyst to the ill-conceived question of "Does it?",
> estimation instead draws the analyst's attention to the question of "How
> much?" --- the very topic that defines quantitative research.

**Estimation statistics** replaces dichotomous significant/non-significant
thinking with a focus on:

- **Effect sizes**: How large is the difference? (mean difference, Cohen's d,
  Cliff's delta, correlation coefficient, regression slope...)
- **Bootstrap confidence intervals**: How precise is our estimate? What is the
  range of plausible values?

This framework, championed by
[Gardner & Altman (1986)](https://doi.org/10.1136/bmj.292.6522.746) and
formalized in the DABEST software
([Ho et al. 2019](https://www.nature.com/articles/s41592-019-0470-3);
[estimationstats.com](https://www.estimationstats.com/#/background)),
produces **estimation plots** that show all individual data points alongside
the effect size and its uncertainty. This gives a much richer picture than a
single p-value:

1. **All data is visible**: Every observation is plotted, exposing the
   distribution shape, outliers, and sample size at a glance.
2. **Effect size is front and center**: The magnitude of the difference is
   displayed directly, not hidden behind a test statistic.
3. **Uncertainty is explicit**: Bootstrap confidence intervals convey
   precision, enabling readers to judge whether the effect is meaningful in
   practice.
4. **Reproducibility**: Effect sizes and CIs travel well across studies and
   facilitate meta-analysis, unlike p-values which depend on sample size.

### A note on p-values

The functions in comparpq do report p-values (permutation test, Welch's
t-test, Mann-Whitney U, or `cor.test`) alongside the estimation statistics.
These are provided **for legacy purposes only**, to satisfy journals that
still require them. Following the recommendations of
[Ho et al. (2019)](https://www.nature.com/articles/s41592-019-0470-3), we
encourage users to:

- **Never** use p-values as the primary basis for interpretation.
- **Focus** on the effect size and its confidence interval.
- State in Methods: *"No null-hypothesis significance testing was performed;
  p-values are provided for legacy purposes only."*

For a thorough introduction to estimation statistics, see
[estimationstats.com/background](https://www.estimationstats.com/#/background).

### Two types of estimation plots

- **Gardner-Altman plot** (2 groups): The raw data and the effect size with
  its bootstrap CI are displayed side by side on a shared axis, making the
  comparison immediately visual.
- **Cumming plot** (3+ groups): All groups are shown on top, and effect sizes
  with CIs are shown below in a separate panel. This layout scales to
  multiple comparisons.

comparpq provides four functions for estimation statistics on phyloseq objects:

| Function | Level | Variable type |
|----------|-------|---------------|
| `estim_diff_pq()` | Single phyloseq | Categorical |
| `estim_cor_pq()` | Single phyloseq | Numeric |
| `estim_diff_lpq()` | list_phyloseq | Categorical |
| `estim_cor_lpq()` | list_phyloseq | Numeric |

## Setup

```{r setup}
library(comparpq)
library(phyloseq)
library(ggplot2)
library(dplyr)
```

```{r data}
data("data_fungi", package = "MiscMetabar")

# Clean data: remove samples with NA in Height
pq <- subset_samples(data_fungi, !is.na(Height))
pq <- clean_pq(pq)
```

## Categorical comparisons with `estim_diff_pq()`

### Two-group comparison (Gardner-Altman plot)

For two groups, `estim_diff_pq()` produces Gardner-Altman plots where the
raw data and the effect size are displayed side by side:

```{r estim_diff_2groups, eval = FALSE}
res <- estim_diff_pq(pq, fact = "Height")

# View summary table
res$summary

# View the Gardner-Altman plot for Hill_0 (richness)
res$plots$Hill_0
```

### Different effect types

You can choose among several effect size measures:

```{r effect_types, eval = FALSE}
# Cohen's d (standardized mean difference)
res_d <- estim_diff_pq(pq, fact = "Height", effect_type = "cohens_d")
res_d$summary

# Hedges' g (bias-corrected Cohen's d for small samples)
res_g <- estim_diff_pq(pq, fact = "Height", effect_type = "hedges_g")
res_g$summary

# Cliff's delta (non-parametric effect size)
res_cliff <- estim_diff_pq(pq, fact = "Height", effect_type = "cliffs_delta")
res_cliff$summary
```

### Three or more groups (Cumming plot)

When the factor has 3+ levels, a Cumming estimation plot is produced
automatically:

```{r estim_diff_3groups, eval = FALSE}
# If your factor has 3+ levels
pq_time <- subset_samples(data_fungi, !is.na(Time))
pq_time <- clean_pq(pq_time)

res_time <- estim_diff_pq(pq_time, fact = "Time")
res_time$plots$Hill_0
```

## Numeric variable analysis with `estim_cor_pq()`

For continuous variables, `estim_cor_pq()` computes bootstrap confidence
intervals for both correlation coefficients and regression slopes:

```{r estim_cor, eval = FALSE}
# Add library size as a numeric variable
sam <- sample_data(pq)
sam$lib_size <- sample_sums(pq)
sample_data(pq) <- sam

res_cor <- estim_cor_pq(pq, variable = "lib_size")

# Correlation estimates with bootstrap CIs
res_cor$correlations

# Regression slopes with bootstrap CIs
res_cor$regressions

# Scatter plot with regression line and CI ribbon
res_cor$plots$Hill_0
```

### Spearman correlation

```{r spearman, eval = FALSE}
res_spearman <- estim_cor_pq(pq, variable = "lib_size", method = "spearman")
res_spearman$correlations
```

## Comparing across pipelines with list_phyloseq

### Categorical: `estim_diff_lpq()`

Apply estimation statistics across multiple phyloseq objects in a
`list_phyloseq`:

```{r estim_diff_lpq, eval = FALSE}
lpq <- list_phyloseq(
  list(
    fungi = pq,
    fungi_copy = pq
  ),
  same_bioinfo_pipeline = FALSE
)

res_lpq <- estim_diff_lpq(lpq, fact = "Height")

# Combined summary across all pipelines
res_lpq$summary

# Access individual results
res_lpq$results$fungi$plots$Hill_0
```

### Numeric: `estim_cor_lpq()`

```{r estim_cor_lpq, eval = FALSE}
# Assuming lib_size is in sample_data of all phyloseq objects
res_cor_lpq <- estim_cor_lpq(lpq, variable = "lib_size")

# Compare correlations across pipelines
res_cor_lpq$correlations

# Compare regression slopes
res_cor_lpq$regressions
```

## Custom diversity functions

You can supply any custom diversity function instead of Hill numbers.
The function must take a phyloseq object and return either:

- A named numeric vector (names = sample names)
- A data.frame with one row per sample

```{r custom_fn, eval = FALSE}
# Example: use Pielou's evenness
my_evenness <- function(physeq) {
  otu <- as.data.frame(otu_table(physeq))
  if (!taxa_are_rows(physeq)) otu <- t(otu)
  otu <- t(otu)
  H <- vegan::diversity(otu, index = "shannon")
  S <- vegan::specnumber(otu)
  J <- H / log(S)
  stats::setNames(J, sample_names(physeq))
}

res_even <- estim_diff_pq(pq, fact = "Height", custom_fn = my_evenness)
res_even$summary
```

## References

- Gardner, M. J., & Altman, D. G. (1986). Confidence intervals rather than P
  values: estimation rather than hypothesis testing. *BMJ*, 292(6522), 746-750.
  <https://doi.org/10.1136/bmj.292.6522.746>
- Ho, J., Tumkaya, T., Aryal, S., Choi, H., & Claridge-Chang, A. (2019).
  Moving beyond P values: data analysis with estimation graphics. *Nature
  Methods*, 16(7), 565-566.
  <https://www.nature.com/articles/s41592-019-0470-3>
- Claridge-Chang, A., & Assam, P. N. (2016). Estimation statistics should
  replace significance testing. *Nature Methods*, 13(2), 108-109.

### Online resources

- [estimationstats.com](https://www.estimationstats.com/#/background) ---
  Interactive web application for estimation statistics with background
  tutorial, free dataset upload, and automated estimation plots.
- [dabestr R package](https://cran.r-project.org/package=dabestr) ---
  The R implementation of DABEST used internally by comparpq.
