---
title: "Getting Started with comparpq"
author: "Adrien Taudière"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with comparpq}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 5
)
```

## Introduction

**comparpq** is an extension package to [{MiscMetabar}](https://adrientaudiere.github.io/MiscMetabar/) and [{phyloseq}](https://joey711.github.io/phyloseq/) designed for comparing phyloseq objects. This package is particularly useful for:

- Evaluating taxonomic assignment accuracy using mock communities
- Creating interactive visualizations of microbiome data
- Comparing multiple taxonomic databases or assignment methods
- Data preprocessing and quality control

## Installation

You can install comparpq from GitHub:

```{r installation, eval=FALSE}
# Install from GitHub
if (!requireNamespace("devtools", quietly = TRUE)) {
  install.packages("devtools")
}
devtools::install_github("adrientaudiere/comparpq")
```

## Loading Required Libraries

```{r setup}
library(comparpq)

# The package automatically loads dependencies including:
# - phyloseq for handling microbiome data
# - MiscMetabar for additional microbiome analysis tools
```

## Working with Example Data

The package includes the `Glom_otu` dataset, which contains glomalean (arbuscular mycorrhizal fungi) OTU data:

```{r load_data}
# Load the example dataset
data("Glom_otu")

# Examine the phyloseq object structure
Glom_otu
```

```{r examine_data}
# Basic information about the dataset
cat("Number of taxa:", phyloseq::ntaxa(Glom_otu), "\n")
cat("Number of samples:", phyloseq::nsamples(Glom_otu), "\n")
cat("Taxonomic ranks:", paste(phyloseq::rank_names(Glom_otu), collapse = ", "), "\n")
```

## Interactive Visualizations

### Bubble Plots with Observable HQ

One of the key features of comparpq is the ability to create interactive bubble plots using Observable HQ:

```{r bubble_basic, eval=FALSE}
# Basic bubble plot
bubbles_pq(Glom_otu)
```

```{r bubble_colored, eval=FALSE}
# Bubble plot colored by Family
bubbles_pq(Glom_otu, 
          rank_color = "Family",
          min_nb_seq = 1000)
```

```{r bubble_custom, eval=FALSE}
# Customized bubble plot with different color scheme
bubbles_pq(Glom_otu, 
          rank_color = "Class",
          categorical_scheme = "d3.schemePastel1",
          min_nb_seq = 500,
          log1ptransform = TRUE)
```

### Traditional Plots

For non-interactive visualizations, you can use:

```{r trad_plots, eval=FALSE}
# Create comparison matrices for visualization
matrix_data <- tc_points_matrix(Glom_otu)

# Bar chart comparison (requires appropriate data structure)
# tc_bar(matrix_data)

# Circular comparison plot
# tc_circle(matrix_data)
```

## Data Manipulation and Preprocessing

### Working with Taxonomic Tables

```{r taxo_manipulation}
# Select specific taxonomic ranks
selected_ranks <- select_ranks_pq(Glom_otu, ranks_names = c("Kingdom", "Phylum", "Class"))
cat("Original ranks:", paste(phyloseq::rank_names(Glom_otu), collapse = ", "), "\n")
cat("Selected ranks:", paste(phyloseq::rank_names(selected_ranks), collapse = ", "), "\n")
```

```{r rename_ranks}
# Rename taxonomic ranks
renamed_physeq <- rename_ranks_pq(Glom_otu, 
                                 new_names = c("Domain", "Division", "Class_level", 
                                             "Order_level", "Family_level", "Genus_level"))
cat("Renamed ranks:", paste(phyloseq::rank_names(renamed_physeq), collapse = ", "), "\n")
```

### Pattern Replacement and Data Cleaning

```{r pattern_replace}
# Replace unwanted patterns with NA
# This is useful for cleaning taxonomic assignments
cleaned_physeq <- taxtab_replace_pattern_by_NA(Glom_otu, 
                                              pattern = "^uncultured", 
                                              ranks_names = "Genus")

# Check how many taxa were affected
original_genus <- phyloseq::tax_table(Glom_otu)[, "Genus"]
cleaned_genus <- phyloseq::tax_table(cleaned_physeq)[, "Genus"]
cat("Uncultured genera replaced with NA:", 
    sum(is.na(cleaned_genus)) - sum(is.na(original_genus)), "\n")
```

## Mock Community Analysis

### Creating Fake Taxa for Accuracy Assessment

To evaluate taxonomic assignment accuracy, you can add fake taxa to your dataset:

```{r fake_taxa}
# Add shuffled sequences (creates fake taxa from existing sequences)
physeq_with_shuffled <- add_shuffle_seq_pq(Glom_otu, nb_seq_fake = 10)
cat("Original taxa count:", phyloseq::ntaxa(Glom_otu), "\n")
cat("With fake taxa:", phyloseq::ntaxa(physeq_with_shuffled), "\n")

# Add external sequences (simulates contamination or database gaps)
physeq_with_external <- add_external_seq_pq(Glom_otu, nb_external_seq = 5)
cat("With external sequences:", phyloseq::ntaxa(physeq_with_external), "\n")
```

### Accuracy Metrics Computation

```{r metrics_example, eval=FALSE}
# Example of computing accuracy metrics
# This requires properly formatted comparison data

# Step 1: Prepare your comparison data structure
# ranks_df should contain your taxonomic assignments to evaluate
# true_values_df should contain the known correct assignments

# Example structure (replace with your actual data):
# ranks_df <- data.frame(
#   method1 = c("Fungi", "Ascomycota", "Sordariomycetes"),
#   method2 = c("Fungi", "Basidiomycota", "Agaricomycetes")
# )
# 
# true_values_df <- data.frame(
#   Kingdom = c("Fungi", "Fungi"),
#   Phylum = c("Glomeromycota", "Glomeromycota")
# )

# Step 2: Compute metrics
# accuracy_results <- tc_metrics_mock(physeq_with_shuffled, 
#                                   ranks_df = ranks_df,
#                                   true_values_df = true_values_df,
#                                   fake_taxa = TRUE)
```

## Visualization of NA Patterns

The package includes specialized visualization for understanding missing taxonomic data:

```{r rainplot, eval=FALSE}
# Visualize NA patterns in taxonomic assignments
rainplot_taxo_na(Glom_otu)
```

## Resolving Taxonomic Conflicts

When working with multiple taxonomic databases, conflicts may arise:

```{r resolve_conflicts, eval=FALSE}
# Resolve taxonomic conflicts between different assignment methods
# This function helps when you have conflicting taxonomic assignments
resolved_physeq <- resolve_taxo_conflict(Glom_otu, 
                                        conflict_rank = "Genus",
                                        resolve_method = "majority")
```

## Advanced Usage Tips

### Workflow for Method Comparison

1. **Prepare your data**: Start with a phyloseq object containing your OTU/ASV data
2. **Add fake taxa**: Use `add_shuffle_seq_pq()` and `add_external_seq_pq()` to create test cases
3. **Perform taxonomic assignments**: Apply different methods/databases to your data
4. **Compare results**: Use `tc_metrics_mock()` to quantify accuracy
5. **Visualize**: Create plots using `bubbles_pq()`, `tc_bar()`, or `tc_circle()`

### Data Quality Control

```{r qc_workflow}
# Example quality control workflow
qc_physeq <- Glom_otu

# 1. Clean up problematic annotations
qc_physeq <- taxtab_replace_pattern_by_NA(qc_physeq, 
                                         pattern = "unknown|unidentified|uncultured", 
                                         ranks_names = c("Family", "Genus"))

# 2. Select relevant taxonomic ranks
qc_physeq <- select_ranks_pq(qc_physeq, 
                            ranks_names = c("Kingdom", "Phylum", "Class", "Order", "Family"))

# 3. Standardize rank names
qc_physeq <- rename_ranks_pq(qc_physeq, 
                            new_names = c("Kingdom", "Phylum", "Class", "Order", "Family"))

cat("Quality control completed. Final dataset has", phyloseq::ntaxa(qc_physeq), "taxa\n")
```

## Getting Help

- **Function documentation**: Use `?function_name` for detailed help on any function
- **Package website**: Visit https://adrientaudiere.github.io/comparpq/
- **Report issues**: https://github.com/adrientaudiere/comparpq/issues
- **Related packages**: 
  - [{MiscMetabar}](https://adrientaudiere.github.io/MiscMetabar/): Extended microbiome analysis
  - [{phyloseq}](https://joey711.github.io/phyloseq/): Core microbiome data handling

## Summary

The comparpq package provides a comprehensive toolkit for:

- **Interactive visualization** of phyloseq objects using modern web technologies
- **Accuracy assessment** of taxonomic assignment methods using mock communities  
- **Data preprocessing** and quality control for microbiome analyses
- **Comparative analysis** between different taxonomic databases or assignment approaches

This makes it particularly valuable for researchers working on taxonomic assignment method development, database comparison, and microbiome data quality assessment.