---
title: "The list_phyloseq Class"
author: "Adrien Taudière"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The list_phyloseq Class}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 5
)
```

## Introduction

The `list_phyloseq` class is an S7 class designed to store and compare multiple phyloseq objects. It automatically computes summary statistics for each phyloseq object and determines the type of comparison based on:

- **Detected characteristics**: sample overlap, nested samples, shared modalities
- **User-provided parameters**: `same_primer_seq_tech` and `same_bioinfo_pipeline`

This vignette demonstrates:

- How to create `list_phyloseq` objects with the appropriate parameters
- The six types of comparisons and when to use them
- Utility functions for manipulating `list_phyloseq` objects
- Visualization functions for exploring and presenting results

## Setup

```{r setup}
devtools::load_all()
# library(comparpq)
library(phyloseq)
```

## Creating a list_phyloseq Object

The simplest way to create a `list_phyloseq` is to pass a named list of phyloseq objects. You can also specify the `same_primer_seq_tech` and `same_bioinfo_pipeline` parameters to control the comparison type:

```{r create_lpq}
lpq <- list_phyloseq(list(
  fungi = data_fungi,
  fungi_mini = data_fungi_mini
))

lpq
```

The output shows:

- **Summary table**: Statistics for each phyloseq object (n_samples, n_taxa, n_sequences, etc.)
- **Comparison characteristics**: Type of comparison, same_primer_seq_tech, same_bioinfo_pipeline, sample overlap, shared modalities

### Key Parameters

When creating a `list_phyloseq`, you can specify:

- `same_primer_seq_tech` (logical, default TRUE): Set to FALSE when comparing different primers (e.g., ITS1 vs ITS2) or technologies (e.g., Illumina vs PacBio)
- `same_bioinfo_pipeline` (logical, default TRUE): Set to FALSE when comparing different clustering methods, taxonomic databases, or analysis parameters

## Types of Comparisons

The `list_phyloseq` class determines the type of comparison based on detected sample overlap and the user-provided `same_primer_seq_tech` and `same_bioinfo_pipeline` parameters. There are six main types:


### Summary of Comparison Types

| Type | Same Samples | same primer/seq_tech | same pipeline  | same pipeline same modality | Use Case | same modality | Use Case |
|------|--------------|----------------------|-----------------------|----------||----------|
| REPRODUCIBILITY | Yes | TRUE | TRUE |  TRUE | Test reproducibility of identical analysis |
| ROBUSTNESS | Yes | TRUE | FALSE |  TRUE | Test effect of different pipelines/parameters |
| REPLICABILITY | Yes | FALSE | - |  TRUE | Test effect of different primers/technologies |
| NESTED_ROBUSTNESS | Nested | - | - |  TRUE | Test effect of rarefaction/filtering |
| EXPLORATION | No | - | - |  TRUE | Explore differences between sample groups |
| SEPARATE_ANALYSIS | No | - | - |  FALSE | Independent analyses (comparison may not be meaningful) |


### 1. REPRODUCIBILITY

When all phyloseq objects have the **same samples**, the **same primer/technology** (`same_primer_seq_tech = TRUE`), and the **same pipeline** (`same_bioinfo_pipeline = TRUE`, both defaults), the comparison is classified as REPRODUCIBILITY.

```{r reproducibility_example}
# REPRODUCIBILITY: Same samples, same pipeline (default parameters)
# Example: Running the same analysis twice to test reproducibility
lpq_repro <- list_phyloseq(list(
  run1 = data_fungi,
  run2 = data_fungi
))

cat("Type:", lpq_repro@comparison$type_of_comparison, "\n")
cat("Same primer/seq tech:", lpq_repro@comparison$same_primer_seq_tech, "\n")
cat("Same bioinfo pipeline:", lpq_repro@comparison$same_bioinfo_pipeline, "\n")
```

This can also be useful to test for effect of stochasticity in some function.

```{r reproducibility_example_rarefy}
# REPRODUCIBILITY: Same samples, same pipeline (default parameters)
# Example: Running the same analysis twice to test reproducibility
lpq_repro <- list_phyloseq(list(
  run1 = rarefy_even_depth(data_fungi, rngseed = 66),
  run2 = rarefy_even_depth(data_fungi, rngseed = 42)
))

cat("Type:", lpq_repro@comparison$type_of_comparison, "\n")
cat("Same primer/seq tech:", lpq_repro@comparison$same_primer_seq_tech, "\n")
cat("Same bioinfo pipeline:", lpq_repro@comparison$same_bioinfo_pipeline, "\n")
```




### 2. ROBUSTNESS

When all phyloseq objects have the **same samples** and the **same primer/technology**, but **different pipelines** (`same_bioinfo_pipeline = FALSE`), the comparison is classified as ROBUSTNESS.

```{r robustness_example}
# ROBUSTNESS: Same samples, different pipeline
# Example: Comparing different clustering methods or taxonomic databases
lpq_robust <- list_phyloseq(
  list(
    raw_fungi = data_fungi,
    fungi_swarm = postcluster_pq(data_fungi, method = "swarm"),
    fungi_mumu = mumu_pq(data_fungi)$new_physeq
  ),
  same_bioinfo_pipeline = FALSE # Different bioinformatics pipelines
)

cat("Type:", lpq_robust@comparison$type_of_comparison, "\n")
cat("Same primer/seq tech:", lpq_robust@comparison$same_primer_seq_tech, "\n")
cat("Same bioinfo pipeline:", lpq_robust@comparison$same_bioinfo_pipeline, "\n")
```

### 3. REPLICABILITY

When all phyloseq objects have the **same samples** but **different primers or technologies** (`same_primer_seq_tech = FALSE`), the comparison is classified as REPLICABILITY.

```{r replicability_example}
# REPLICABILITY: Same samples, different primer/technology
# Example: Comparing ITS1 vs ITS2, or Illumina vs PacBio
lpq_replic <- list_phyloseq(
  list(
    ITS1 = data_fungi,
    ITS2 = data_fungi # In real use, this would be data from a different primer
  ),
  same_primer_seq_tech = FALSE # Different primers/technologies
)

cat("Type:", lpq_replic@comparison$type_of_comparison, "\n")
cat("Same primer/seq tech:", lpq_replic@comparison$same_primer_seq_tech, "\n")
```

### 4. NESTED_ROBUSTNESS

When one phyloseq object is **derived from another** (e.g., rarefied version), the samples from one are a subset of samples from another. This type is detected automatically regardless of the `same_primer_seq_tech` and `same_bioinfo_pipeline` parameters. It is useful for testing robustness to data processing choices like rarefaction or filtering.

```{r nested_example}
# Create a rarefied version (samples may be lost if they don't meet depth threshold)
set.seed(123)
data_fungi_rarefied <- rarefy_even_depth(data_fungi, sample.size = 1000, verbose = FALSE)

lpq_nested <- list_phyloseq(list(
  original = data_fungi,
  rarefied = data_fungi_rarefied
))

# Check the comparison type
cat("Type:", lpq_nested@comparison$type_of_comparison, "\n")
cat("Nested samples:", lpq_nested@comparison$nested_samples, "\n")
cat("Nesting structure:", lpq_nested@comparison$nesting_structure, "\n")
```

The comparison should focus on the **common (nested) samples** to make meaningful comparisons. Use `filter_common_lpq()` for this (see [Utility Functions](#utility-functions)).

### 5. EXPLORATION

When phyloseq objects have **different samples but shared modalities** in sample_data (e.g., same experimental conditions, same sites), the comparison is classified as EXPLORATION. This type is detected automatically based on sample_data.

```{r exploration_example}
# Simulate by subsetting samples from different parts of the dataset
# In real use, these would be different sample groups with shared experimental factors
sample_names_fungi <- sample_names(data_fungi)
n_samples <- length(sample_names_fungi)

# Split samples into two groups (simulating different sample sets with shared modalities)
group1_samples <- sample_names_fungi[1:floor(n_samples / 2)]
group2_samples <- sample_names_fungi[(floor(n_samples / 2) + 1):n_samples]

data_group1 <- prune_samples(group1_samples, data_fungi)
data_group2 <- prune_samples(group2_samples, data_fungi)

lpq_exploration <- list_phyloseq(list(
  group1 = data_group1,
  group2 = data_group2
))

cat("Type:", lpq_exploration@comparison$type_of_comparison, "\n")
cat("Same samples:", lpq_exploration@comparison$same_samples, "\n")
cat("Common samples:", lpq_exploration@comparison$n_common_samples, "\n")
```

### 6. SEPARATE_ANALYSIS

When phyloseq objects have **different samples with no shared modalities**, the comparison is classified as SEPARATE_ANALYSIS. Direct comparison may not be meaningful. Separate analysis of each phyloseq object is recommended.

```{r separate_example}
# Example using completely different datasets
data("enterotype", package = "phyloseq")

lpq_separate <- list_phyloseq(list(
  fungi = data_fungi,
  enterotype = enterotype
))

cat("Type:", lpq_separate@comparison$type_of_comparison, "\n")
cat("Same samples:", lpq_separate@comparison$same_samples, "\n")
```


## Utility Functions {#utility-functions}

### Adding and Removing phyloseq Objects

```{r add_remove}
# Start with a single phyloseq
lpq_single <- list_phyloseq(list(fungi = data_fungi))
cat("Initial number of phyloseq objects:", length(lpq_single), "\n")

# Add another phyloseq object
lpq_added <- add_phyloseq(lpq_single, data_fungi_mini, name = "fungi_mini")
cat("After adding:", length(lpq_added), "\n")
cat("Names:", paste(names(lpq_added), collapse = ", "), "\n")

# Remove a phyloseq object
lpq_removed <- remove_phyloseq(lpq_added, "fungi_mini")
cat("After removing:", length(lpq_removed), "\n")
```

### Updating Summary and Comparison

If you modify the phyloseq objects inside the list_phyloseq, use `update_list_phyloseq()` to recompute the summary table and comparison characteristics. You can also use this function to change the `same_primer_seq_tech` or `same_bioinfo_pipeline` parameters:

```{r update_lpq, eval=FALSE}
lpq@phyloseq_list[[1]] <- rarefy_even_depth(lpq@phyloseq_list[[1]])
lpq_updated <- update_list_phyloseq(lpq)

# Change the comparison type by modifying parameters
lpq_robustness <- update_list_phyloseq(lpq, same_bioinfo_pipeline = FALSE)
cat("New type:", lpq_robustness@comparison$type_of_comparison, "\n")
```

### Filtering to Common Samples and Taxa

The `filter_common_lpq()` function filters each phyloseq object to retain only samples and/or taxa that are common across all objects. This is particularly useful for:

- **NESTED_ROBUSTNESS** comparisons: filter to common samples when comparing original vs rarefied
- Any comparison where you need all phyloseq objects to have the same samples/taxa

```{r filter_common}
# Create a list_phyloseq with nested samples
set.seed(123)
lpq_to_filter <- list_phyloseq(list(
  original = data_fungi,
  rarefied = rarefy_even_depth(data_fungi, sample.size = 1000, verbose = FALSE)
))

# Before filtering
cat("Before filtering:\n")
cat("  Original samples:", nsamples(lpq_to_filter[["original"]]), "\n")
cat("  Rarefied samples:", nsamples(lpq_to_filter[["rarefied"]]), "\n")
cat("  Common samples:", lpq_to_filter@comparison$n_common_samples, "\n")

# Filter to keep only common samples
lpq_filtered <- filter_common_lpq(lpq_to_filter,
  filter_samples = TRUE,
  filter_taxa = FALSE,
  verbose = TRUE
)

# After filtering - both have the same samples
cat("\nAfter filtering:\n")
cat("  Original samples:", nsamples(lpq_filtered[["original"]]), "\n")
cat("  Rarefied samples:", nsamples(lpq_filtered[["rarefied"]]), "\n")
cat("  Same samples now:", lpq_filtered@comparison$same_samples, "\n")
```

You can also filter to common taxa:

```{r filter_taxa, eval=FALSE}
# Filter both samples and taxa
lpq_filtered_both <- filter_common_lpq(lpq_to_filter,
  filter_samples = TRUE,
  filter_taxa = TRUE
)
```

### Viewing Shared Modalities

The `shared_mod_lpq()` function displays which sample_data variable modalities are shared across phyloseq objects:

```{r shared_mod}
shared_mod_lpq(lpq)
```

This is useful for understanding what factors are common across your datasets, especially for EXPLORATION type comparisons.

## Visualization Functions

### formattable_lpq: Summary Table Visualization

The `formattable_lpq()` function creates a beautiful HTML table showing the summary statistics with colored bars and indicators:

```{r formattable, eval=FALSE}
# Basic formattable visualization
formattable_lpq(lpq)

# Custom columns selection
formattable_lpq(lpq, columns = c("name", "n_samples", "n_taxa", "n_sequences"))

# Custom colors for bars
formattable_lpq(lpq, bar_colors = list(
  n_samples = "steelblue",
  n_taxa = "darkgreen"
))
```

For a more complete view including comparison characteristics:

```{r formattable_full, eval=FALSE}
# Get both summary and comparison tables
result <- formattable_lpq_full(lpq)
result$summary
result$comparison
```

### upset_lpq: Set Intersection Visualization

The `upset_lpq()` function creates UpSet plots or Venn diagrams showing shared taxonomic values across phyloseq objects:

```{r upset_venn, fig.height=4}
# Create a list_phyloseq with multiple objects
lpq_multi <- list_phyloseq(list(
  fungi = data_fungi,
  fungi_mini = data_fungi_mini
))

# Venn diagram (automatically chosen for ≤4 sets)
upset_lpq(lpq_multi, tax_rank = "Family")
```

```{r upset_plot, fig.height=6, eval=FALSE}
# Force UpSet plot (better for more than 4 sets or complex intersections)
data("enterotype", package = "phyloseq")
lpq_many <- list_phyloseq(list(
  fungi = data_fungi,
  fungi_mini = data_fungi_mini,
  fungi_rarefied = rarefy_even_depth(data_fungi, sample.size = 1000, verbose = FALSE),
  enterotype = enterotype
))

upset_lpq(lpq_many, tax_rank = "Genus", plot_type = "upset")
```

## Subsetting and Accessing Elements

The `list_phyloseq` class supports standard R subsetting operations:

```{r subsetting}
# Get the number of phyloseq objects
length(lpq)

# Get names
names(lpq)

# Extract a single phyloseq object
pq1 <- lpq[["fungi"]]
cat("Extracted phyloseq has", ntaxa(pq1), "taxa\n")

# Subset to create a new list_phyloseq with fewer objects
lpq_subset <- lpq[1] # Returns a new list_phyloseq with only the first object
```

## Workflow Example: Comparing Rarefaction Effect

Here's a complete workflow for comparing the effect of rarefaction on your data:

```{r workflow, fig.height=5}
# Step 1: Create original and rarefied versions
set.seed(42)
original <- data_fungi
rarefied <- rarefy_even_depth(data_fungi, sample.size = 1000, verbose = FALSE)

# Step 2: Create list_phyloseq
lpq_rarefaction <- list_phyloseq(list(
  original = original,
  rarefied = rarefied
))

# Step 3: Check the comparison type
print(lpq_rarefaction)

# Step 4: Filter to common samples for fair comparison
lpq_common <- filter_common_lpq(lpq_rarefaction,
  filter_samples = TRUE,
  verbose = TRUE
)

# Step 5: Visualize shared taxa at different ranks
upset_lpq(lpq_common, tax_rank = "Family")
```

```{r workflow_summary, eval=FALSE}
# Step 6: View the formatted summary
formattable_lpq(lpq_common)
```

## Summary

The `list_phyloseq` class provides a powerful framework for comparing multiple phyloseq objects:

- **Precise comparison type** determined by `same_primer_seq_tech` and `same_bioinfo_pipeline` parameters combined with automatic detection of sample overlap
- **Summary statistics** computed for each phyloseq object
- **Utility functions** for adding, removing, filtering, and updating
- **Visualization functions** for exploring differences (formattable tables, UpSet/Venn plots)

Key points:

1. Set `same_bioinfo_pipeline = FALSE` for ROBUSTNESS comparisons (different pipelines)
2. Set `same_primer_seq_tech = FALSE` for REPLICABILITY comparisons (different primers/technologies)
3. NESTED_ROBUSTNESS, EXPLORATION, and SEPARATE_ANALYSIS are detected automatically
4. For NESTED_ROBUSTNESS comparisons, use `filter_common_lpq()` to focus on common samples
5. The `upset_lpq()` function is excellent for visualizing shared taxonomic composition
6. Use `update_list_phyloseq()` to change parameters or recompute after modifications

## Getting Help

- **Function documentation**: Use `?function_name` for detailed help
- **Package website**: https://adrientaudiere.github.io/comparpq/
- **Report issues**: https://github.com/adrientaudiere/comparpq/issues
